---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `R`/`uniCATE`

> Univariate Conditional Average Treatment Effect Estimation

__Author:__ [Philippe Boileau](https://pboileau.ca/)

<!-- badges: start -->
[![Check](https://github.com/insightsengineering/uniCATE/actions/workflows/check.yaml/badge.svg)](https://github.com/insightsengineering/uniCATE/actions/workflows/check.yaml)
[![Docs](https://github.com/insightsengineering/uniCATE/actions/workflows/docs.yaml/badge.svg)](https://github.com/insightsengineering/uniCATE/actions/workflows/docs.yaml)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

---

`uniCATE` implements statistical inference procedures for variable importance
measures that assess the treatment effect modification capabilities of
individual pre-treatment biomarkers in high-dimensional randomized control
trials. This variable importance measure is defined as the vector of simple
linear regression slope coefficients obtained by regressing the difference
in potential outcomes on each biomarker. This parameter, which we dub the
*univariate conditional average treatment effect*, is a reasonable indicator of
treatment effect modification in all but pathological biomarker-outcome
relationships, and can therefore be used to identify predictive biomarkers.
Assumption-lean estimation and testing procedures based on semiparametric theory
are made available for continuous, binary, and right-censored time-to-event
outcomes.


## Installation

The _development version_ of the package may be installed from GitHub using
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval=FALSE}
remotes::install_github("insightsengineering/uniCATE")
```

## Usage

`unicate()` should be used when the outcome is continuous or binary. For
right-censored time-to-event outcomes, use `sunicate()`.

## Example

We simulate a randomized control trial in which there is a heterogeneous
treatment effect for biomarkers 1 and 2. `unicate()` successfully identifies
these biomarkers as effect modifiers.

```{r setup, include=FALSE}
library(uniCATE)
library(dplyr)
library(sl3)
```
```{r fake-setup, eval=FALSE}
# load the required libraries
library(uniCATE)
library(dplyr)
library(sl3)
```
```{r example}
# set the seed for reproducibility
set.seed(514)

# simulate some randomized control data
n <- 100
data <- tibble("treatment" = rbinom(n, 1, 0.5)) %>%
  mutate(
    bio1 = rnorm(n, mean = 2, sd = 0.2),
    bio2 = rnorm(n, mean = -2, sd = 0.2),
    bio3 = rnorm(n, mean = 0, sd = 0.1),
    bio4 = rnorm(n, mean = 0, sd = 0.1),
    covar = 0.2 * rbinom(n, 1, 0.4),
    response = covar + bio1 * treatment + bio2 * treatment
  )

# define the required arguments
covariates <- c("bio1", "bio2", "bio3", "bio4", "covar")
biomarkers <- c("bio1", "bio2", "bio3", "bio4")
propensity_score_ls <- list("1" = 0.5, "0" = 0.5)

# create a simple SuperLearner using a linear model and a random forest
interactions <- lapply(biomarkers, function(b) c(b, "treatment"))
lrnr_interactions <- sl3::Lrnr_define_interactions$new(interactions)
lrnr_glm <- sl3::make_learner(
  sl3::Pipeline, lrnr_interactions, sl3::Lrnr_glm$new()
)
lrnr_sl <- Lrnr_sl$new(
  learners = make_learner(
    Stack, Lrnr_ranger$new(), lrnr_glm
  ),
  metalearner = make_learner(Lrnr_nnls)
)

# apply uniCATE to the simulated data
unicate(
  data,
  outcome = "response",
  treatment = "treatment",
  covariates = covariates,
  biomarkers = biomarkers,
  propensity_score_ls = propensity_score_ls,
  super_learner = lrnr_sl,
  v_folds = 2L
)
```

## Issues

If you encounter any bugs or have any specific feature requests, please [file
an issue](https://github.com/insightsengineering/uniCATE/issues).


## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/insightsengineering/uniCATE/blob/master/.github/CONTRIBUTING.md)
prior to submitting a pull request.


## License

The contents of this repository are distributed under the MIT license. See file
[`LICENSE.md`](https://github.com/insightsengineering/uniCATE/blob/main/LICENSE.md)
for details.

*Note: This license is a placeholder until the legal team is consulted.*
