---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `R`/`uniCATE`

> Univarite Conditional Average Treatment Effect Estimation

__Author:__ [Philippe Boileau](https://pboileau.ca/)

<!-- badges: start -->
<!-- badges: end -->

---

`uniCATE` implements an efficient semiparametric procedure for the estimation of
individual variables' conditional average treatment effects' best linear
approximations. This method can identify potential predictive biomarkers in
randomized and observational studies using assumption-lean statistical inference
techniques that rely on flexible machine learning algorithms.


## Installation

The _development version_ of the package may be installed from GitHub using
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("insightsengineering/uniCATE")
```
