---
title: "Using uniCATE"
author: "[Philippe Boileau](https://pboileau.ca)"
date: "`r Sys.Date()`"
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using uniCATE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(uniCATE)
library(dplyr)
```


# Background

Predictive biomarkers are biometric measurements that subdivide patient
populations into groups which draw differing benefits from a given treatment.
These biomarkers therefore play an important role in precision medicine: as
indicators of treatment effect modification, they can inform patient treatment
decisions. Predictive biomarkers may also inform drug target discovery, leading
to an improved understanding of novel medications' functioning.

The discovery of predictive biomarkers through statistical techniques has, to
date, largely been a byproduct of optimal treatment rule estimation. As implied
by its name, this tasks consists of learning which treatment would most benefit
a patient given their characteristics. These characteristics could include, but
are not limited to, potentially predictive biomarkers. When interpretable
statistical learning methods are used to estimate these rules, biomarkers found
to be "important" might be declared predictive. For example, estimation methods
based on regularized linear regression might define a biomarker as predictive if
its linear regression coefficient estimate is non-zero [@tian2014].

While this approach to predictive biomarker discovery works well in traditional
asymptotic settings, it is not so when the number of biomarkers is similar to
or larger than the number of patients in a study. These methods offer no
guarantee of false positive rate control [@tian2014;@chen2017], which can have
severe consequences in practice. The inclusion of false positive biomarkers in
diagnostic assays decreases their clinical utility, and already limited
resources are wasted on biological validation experiments of duds in drug target
discovery applications. Patient outcomes are ultimately negatively effected.

We proposed, in recent work, a variable importance parameter that directly
measures individual biomarkers' roles in treatment effect modification. The
methodology, dubbed *uniCATE*, relies on semiparametric theory to perform
assumption-lean inference about this parameter. We demonstrated that our
procedure controls the rate of false discoveries in high-dimensional randomized
control trials, allowing for the accurate identification of predictive
biomarkers. This methodology is implemented in the `uniCATE` package. 


# `uniCATE` in Action

This section provides a high-level overview of the statistical methodology for
various outcome types, and includes brief tutorials of the package's methods
applied to simulated data.

## Continuous Outcomes

### Setup

Suppose we perform a randomized control trial (RCT) where each patient is
represented by one of $n$ independent and identically distributed (i.i.d.)
copies of the full-data random vector $X_i = (W_i, A_i, Y_i^{(0)}, Y_i^{(1)})$
distributed according $P_X$, $i = 1, \ldots, n$. Here, $W_i = (V_i, B_i)$ is a
$(q+p)$-length random vector of $q$ pre-treatment covariates, $V$, like location
and nationality, and $p$ pre-treatment biomarkers, $B$, such as gene expression
data, $A$ is a binary random variable representing treatment assignment, and
$Y^{(1)}$ and $Y^{(0)}$ are continuous random variables corresponding to the
potential outcomes of clinical interest under both treatment and control
conditions, respectively. Of course, we do not have access to patients'
full-data random vector since we only get to observe the outcomes associated
with their treatment assignments. These $n$ i.i.d. random observations are
represented by $O_i = (W_i, A_i, Y_i)$ distributed according to the unknown
data-generating distribution $P_0$. $W_i$ and $A_i$ are defined as before, and
$Y_i = Y_i^{(0)}(1-A_i) + Y^{(1)}_iA_i$. Throughout the rest of the tutorial,
we drop the index $i$ where possible to simplify notation.

### Target Parameter

Ideally, we would like to estimate the full-data parameter
$\Psi(P_x) = (\Psi_1(P_X), \ldots, \Psi_p(P_X)$ corresponding to $p$-length
random vector whose elements are defined as
$$
\Psi_j(P_X) = 
\frac{\mathbb{E}_{P_X}\left[\left(Y^{(1)}-Y^{(0)}\right)B_j\right]}
{\mathbb{E}_{P_X}\left[B_j^2\right]}
$$
for $j=1, \ldots, p$ where we assume that the biomarkers are centered at zero.
Under the assumption that the mean difference in potential outcomes is a linear
function of $B_j$, $\Psi(P_X)$ is the vector of expected simple linear
regression coefficients produced by regressing the difference in potential
outcomes against each biomarker. While the true relationship between the
difference of potential outcomes and a predictive biomarker is almost surely
nonlinear, $\Psi(P_X)$ is a generally informative target of inference. These
parameters are reasonable indicators of all but pathological relationships
between biomarkers and potential outcome differences.

Under conditions that are generally satisfied in the clinical trial setting, we
find that $\Psi(P_X)$ can be estimated from realizations of the observed
data-generating distribution $P_0$. Further, the estimator of this parameter
implemented in the `unicate()` function is asymptotically linear, permitting
valid statistical inference in most RCT scenarios. Details are provided in the
manuscript.

### Example

We simulate some RCT data to demonstrate the application of `unicate()`. We
refrain from generating a high-dimensional vector of biomarkers for
computational convenience.

```{r load-pkgs, eval=FALSE}
library(dplyr)
library(uniCATE)
```
```{r cont-sim}
# simulate n trial participants
n <- 200

# simulate randomized control trial data
set.seed(514)
data <- tibble(
    B1 = rnorm(n, mean = 2, sd = 0.2),
    B2 = rnorm(n, mean = -2, sd = 0.2),
    B3 = rnorm(n, mean = 0, sd = 0.1),
    B4 = rnorm(n, mean = 0, sd = 0.1),
    V1 = rbinom(n, 1, 0.4),
    V2 = rnorm(n),
    A = rbinom(n, 1, 0.5),
    Y = A + V1 - V2 + 2 * B1 * A + B2 * A
)

# define the propensity scores
propensity_score_ls <- list("1" = 0.5, "0" = 0.5)
```

This simulated trial is made up of 200 participants. Four pre-treatment
biomarkers and two pre-treatment covariates are generated for each participant.
Observations are then randomly assigned with equal probability to either the
treatment or the control condition. From the definition of the outcome, it is
clear that `B1` and `B2` are predictive biomarkers. Finally, we define the
`propensity_score_ls`, a list indicating the probability that any given patient
is assigned to the treatment condition, `"1"`, or the control condition, `"0"`.

We now apply `unicate()` to this data to recover the variable importance
measures of the biomarkers.

```{r unicate-cont}
unicate(
  data,
  outcome = "Y",
  treatment = "A",
  covariates = c("B1", "B2", "B3", "B4", "V1", "V2"),
  biomarkers =  c("B1", "B2", "B3", "B4"),
  propensity_score_ls = propensity_score_ls
)
```

The output is a table summarizing the variable importance parameter estimates
of each biomarker (`coef`), the estimators' standard errors (`se`), the
two-sided hypothesis tests' Z-scores (`z`), and the tests' (adjusted) p-values
(`p_value`, `p_value_bh`, and `p_value_holm`).

As expected, `B1` modifies the treatment effect the most, followed by `B2`. Both
of their variable importance measures are found to be significantly different
from zero (using a 5% significance threshold) at the nominal level and after
adjusting for multiple testing. As patients' `B1` and `B2` values marginally
increase, so too does their benefit from being assigned to the treatment
condition over the control condition. The remaining biomarkers' variable
importance parameters are not significantly different from zero at the same
level of significance. There is insufficient evidence to categorize them as
predictive.


## Binary Outcomes

### Setup

The assumed full-data and observed data-generating processes are identical to
those of the continuous outcome, save that $Y^{(0)}, Y^{(1)}$ and $Y$ are binary
random variables.

### Target Causal Parameter

The full-data parameter of interest is slightly modified from the continuous
outcome scenario to reflect the use of a binary outcome. We instead target
$\Psi(P_X) = (\Psi_1(P_X), \ldots, \Psi_p(P_X))$ where
$$
\Psi_j(P_X) =
\frac{\mathbb{E}_{P_X}\left[\left(\mathbb{P}_{P_X}(Y^{(1)} =1| W)
  - \mathbb{P}_{P_X}(Y^{(0)} = 1| W)\right)B_j\right]}
{\mathbb{E}_{P_X}\left[B_j^2\right]}
$$
for centered biomarkers indexed by $j = 1, \ldots, p$.  Assuming a linear
relationship between the difference of the potential outcomes' probability of
success and the covariates, this variable importance parameter consists of the
simple linear regression coefficients of the difference in the conditional
potential outcome success probabilities regressed on each biomarker. While the
true relationship between potential outcome probabilities and covariates is
unlikely to be linear, this parameter is telling of biomarkers' predictive
capacities. Biomarkers with the largest absolute values will correspond to those
that meaningfully modify the treatment effect.

As in the continuous case, this parameter can generally be estimated from
the observed RCT data under non-stringent assumptions about the data-generating
process. Inference is in fact based on the same estimator used for continuous
outcomes, and so we again use `unicate()` to learn about the biomarkers'
marginal importance in regards to binary treatment effect modification. For
details, please review the manuscript.


## Survival Outcomes

### Setup

### Target Causal Parameter

### Example
