---
title: "Using uniCATE"
author: "[Philippe Boileau](https://pboileau.ca)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using uniCATE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uniCATE)
```

# Summary

`uniCATE` implements statistical inference procedures for variable importance
measures that assess the treatment effect modification capabilities of
individual pre-treatment biomarkers in high-dimensional randomized control
trials. These parameters, which we dub the
*univariate conditional average treatment effects*, are reasonable indicators of
treatment effect modification in all but pathological biomarker-outcome
relationships, and can therefore be used to identify predictive biomarkers.
Assumption-lean estimation and testing procedures based on semiparametric theory
are made available for continuous, binary, and right-censored time-to-event
outcomes.

# Statement of Need

Modern semiparametric methods rely on complex statistical theory that is beyond
the training afforded to many of the clinical scientists most interested in
their use. These methods also rely on modern supervised machine learning
algorithms and data-adaptive techniques, like cross-validation, to produce
assumption-lean inference, further raising the barrier for adoption. The method
proposed in our manuscript is no different. The `uniCATE` package therefore
provides a rigorously-tested, open-source implementation of this methodology
that is easy to use.


# `uniCATE` in Action

The `uniCATE` package implements statistical inference procedures targeting
biomarkers' univariate conditional average treatment effects for a range of
outcome types. The following subsections detail the assumed data-generating
processes for the supported outcome classes, their target parameters, and
examples on simulated data.

## Continuous Outcomes

### Setup

Define the full-data random vector. Define the observed data random vector.

### Target Parameter

Define the causal parameter, and then the statistical parameter. In an RCT,
generally identified. See papers for full details and discussion.

### Example

Simulated example from README.

## Binary Outcomes

### Setup

### Target Causal Parameter

### Example

## Survival Outcomes

### Setup

### Target Causal Parameter

### Example
