---
title: "Using uniCATE"
author: "[Philippe Boileau](https://pboileau.ca)"
date: "`r Sys.Date()`"
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using uniCATE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(uniCATE)
```

# Background

Predictive biomarkers are biometric measurements that subdivide patient
populations into groups which draw differing benefits from a given treatment.
These biomarkers therefore play an important role in precision medicine: as
indicators of treatment effect modification, they can inform patient treatment
decisions. Predictive biomarkers may also inform drug target discovery, leading
to an improved understanding of novel medications' functioning.

The discovery of predictive biomarkers through statistical techniques has, to
date, largely been a byproduct of optimal treatment rule estimation. As implied
by its name, this tasks consists of learning which treatment would most benefit
a patient given their characteristics. These characteristics could include, but
are not limited to, potentially predictive biomarkers. When interpretable
statistical learning methods are used to estimate these rules, biomarkers found
to be "important" might be declared predictive. For example, estimation methods
based on regularized linear regression might define a biomarker as predictive if
its linear regression coefficient estimate is non-zero [@tian2014].

While this approach to predictive biomarker discovery works well in traditional
asymptotic settings, it is not so when the number of biomarkers is similar to
or larger than the number of patients in a study. These methods offer no
guarantee of false positive rate control [@tian2014;@chen2017], which can have
severe consequences in practice. The inclusion of false positive biomarkers in
diagnostic assays decreases their clinical utility, and already limited
resources are wasted on biological validation experiments of duds in drug target
discovery applications. Patient outcomes are ultimately negatively effected.

We proposed, in recent work, a variable importance parameter that directly
measures individual biomarkers' roles in treatment effect modification. The
methodology, dubbed *uniCATE*, relies on semiparametric theory to perform
assumption-lean inference about this parameter. We demonstrated that our
procedure controls the rate of false discoveries in high-dimensional randomized
control trials, allowing for the accurate identification of predictive
biomarkers. This methodology is implemented in the `uniCATE` package. 


# `uniCATE` in Action

This section provides a high-level overview of the statistical methodology for
various outcome types, and includes brief tutorials of the package's methods
applied to simulated data.

## Continuous Outcomes

### Setup

Define the full-data random vector. Define the observed data random vector.

### Target Parameter

Define the causal parameter, and then the statistical parameter. In an RCT,
generally identified. See papers for full details and discussion.

### Example

Simulated example from README.

## Binary Outcomes

### Setup

### Target Causal Parameter

### Example

## Survival Outcomes

### Setup

### Target Causal Parameter

### Example
